Classless Java: Tuning Java Interfaces

### 2015-12-06 paper/sections/CaseStudy.tex
review 6 (boh, is ok? non convincing but I do not know what to improve) and exp problem
Review 7-start of 8
revised S7,8,9,10

\section{Case Studies}
In this section we conduct three case studies which reveal various
advantages of using the \mixin annotation. The first case study provides a
simple way to solve the Expression Problem, while supporting multiple,
independent extensions in Java.  The second case study show how to model an
embedded DSL for SQL languages with fluent interfaces. Finally, the third 
case study models a simple game, and compares our implementation with an 
existing one, showing that the amount of code is reduced significantly using \mixin.

@@

The \emph{Expression Problem} (EP)~\cite{wadler98expression} is a well-known
problem about modular extensibility issues in software evolution. Recently, a
new solution~\cite{eptrivially} using only covariant type refinement was
proposed. When this solution is modeled with interfaces and default methods, it
can even provide independent extensibility~\cite{zenger05independentlyextensible}: the ability to assemble a system
from multiple, independently developed extensions. Unfortunatelly, the
required instantiation code makes a plain Java solution verbose and cumbersome
to use. The \mixin annotation is enough to remove the boilerplate code, making
the presented approach very appealing.

@@

%\marco{text under need to be improved}
\paragraph{Adding a New Operation} The difficulty of the EP in OO  
languages arises from adding new operations. For example, adding a pretty printing 
operation would typically change all existing code. However, a solution 
%to the EP forbids this and it also forbids using casts. In other words, it should be possible 
should add the pretty printing operation in a type-safe and modular way.  It
turns out to be easily achieved with the assitance of \mixin.  The code in
Figure~\ref{fig:ep} (rhs) shows how to add the new operation \texttt{print}.
Interface \texttt{ExpP} extends \texttt{Exp} with the extra method
\texttt{print()}. Interfaces \texttt{LitP} and \texttt{AddP} are defined with
default implementations of \texttt{print()}, extending base interfaces
\texttt{Lit} and \texttt{Add}, respectively. Importantly, note that in
\texttt{AddP}, the types of ``\emph{fields}'' (i.e. the getter methods)
\texttt{e1} and \texttt{e2} are refined. If the types were not refined then
\texttt{print()} method in \texttt{AddP} would fail to type-check.

@@

\noindent \texttt{ExpPC} is the new expression interface supporting
\texttt{print} and \texttt{collectLit} operations; \texttt{LitPC} and
\texttt{AddPC} are the extended variants. Notice that except for the routine of
\textbf{extends} clauses, no glue code is required. Return types of
\texttt{e1,e2} are also automatically refined to \texttt{ExpPC}.

Note that the code for instantiation is automatically generated by \mixin. 
Creating a simple expression of type \texttt{ExpPC} is 
as simple as:
\begin{lstlisting}
ExpPC e8 = AddPC.of(LitPC.of(3), LitPC.of(4));
\end{lstlisting}
\noindent In contrast, in a pure Java solution, the tedious instantiation code would need 
to be defined manually.

@@

\lstinputlisting[linerange=37-37]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_QUERY2

>>>>

\section{Case Studies}
In this section we conduct three case studies which reveal various
advantages of using the \mixin annotation. The first case study provides a
simple way to solve the Expression Problem, while supporting multiple,
independent extensions in Java.  The second case study shows how to model an
embedded DSL for SQL languages with fluent interfaces. Finally, the third 
case study models a simple game, and compares our implementation with an 
existing one, showing that the amount of code is reduced significantly using \mixin.

@@

The \emph{Expression Problem} (EP)~\cite{wadler98expression} is a well-known
problem about modular extensibility issues in software evolution. Recently, a
new solution~\cite{eptrivially} using only covariant type refinement was
proposed. When this solution is modeled with interfaces and default methods, it
can even provide independent extensibility~\cite{zenger05independentlyextensible}: the ability to assemble a system
from multiple, independently developed extensions. Unfortunately, the
required instantiation code makes a plain Java solution verbose and cumbersome
to use. The \mixin annotation is enough to remove the boilerplate code, making
the presented approach very appealing.

@@

%\marco{text under need to be improved}
\paragraph{Adding a New Operation} The difficulty of the EP in OO  
languages arises from adding new operations. For example, adding a pretty printing 
operation would typically change all existing code. However, a solution 
%to the EP forbids this and it also forbids using casts. In other words, it should be possible 
should add the pretty printing operation in a type-safe and modular way.  It
turns out to be easily achieved with the assistance of \mixin.  The code in
Figure~\ref{fig:ep} (rhs) shows how to add the new operation \texttt{print}.
Interface \texttt{ExpP} extends \texttt{Exp} with the extra method
\texttt{print()}. Interfaces \texttt{LitP} and \texttt{AddP} are defined with
default implementations of \texttt{print()}, extending base interfaces
\texttt{Lit} and \texttt{Add}, respectively. Importantly, note that in
\texttt{AddP}, the types of ``\emph{fields}'' (i.e. the getter methods)
\texttt{e1} and \texttt{e2} are refined. If the types were not refined then
\texttt{print()} method in \texttt{AddP} would fail to type-check.

@@

\noindent \texttt{ExpPC} is the new expression interface supporting
\texttt{print} and \texttt{collectLit} operations; \texttt{LitPC} and
\texttt{AddPC} are the extended variants. Notice that except for the routine of
\textbf{extends} clauses, no glue code is required. Return types of
\texttt{e1,e2} must be refined to \texttt{ExpPC}.

Note that the code for instantiation is automatically generated by \mixin. 
Creating a simple expression of type \texttt{ExpPC} is 
as simple as:
\begin{lstlisting}
ExpPC e8 = AddPC.of(LitPC.of(3), LitPC.of(4));
\end{lstlisting}
\noindent Without our approach, tedious instantiation code would need 
to be defined manually.

@@

\lstinputlisting[linerange=37-37]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_QUERY2
\marco{put on two lines}

### 2015-12-06 paper/sections/Conclusion.tex
revised S7,8,9,10

 More importantly, we identify the concept of \emph{language tuning}.
 We identify libraries that are already performing language tuning (Lombock and Cofoja), and 
 we forecast many different kinds of language tuning will emerge on suitable platforms like Java or the C\# CLR.
 We identify varius kinds of safety guarancees that can be offered by language tuning, but the door is open for more flavours of safety guarantees to emerge.

>>>>

 More importantly, we identify the concept of \emph{language tuning}.
 We identify libraries that are already performing language tuning (Lombok and Cofoja), and 
 we forecast many different kinds of language tuning will emerge on suitable platforms like Java or the C\# CLR.
 We identify various kinds of safety guarantees that can be offered by language tuning, but the door is open for more flavours of safety guarantees to emerge.

### 2015-12-06 paper/sections/FutureWork.tex
First round over.
Do we cite cofoja ?
revised S7,8,9,10

We are unsure about the value of this workaround because of its trickiness. In
order to define a state with initialization, users have to defining a method trusting that it will be overridden later with a behavioural change, but
such change is visible only after the first time the setter is called. Moreover, this code would cache
the result instead of re-computing it every time. This can be very tricky in a
non functional setting.
%\bruno{Yanlin please polish text/break long sentences.}
\end{comment}

@@

\begin{lstlisting}
interface Point{ ...
  public static Point of(int _x, int _y){
    return new Point(){...
      public Point clone(){ return Point.super.clone();}};}}
\end{lstlisting}

>>>>

We are unsure about the value of this workaround because of its trickiness. In
order to define a state with initialization, users have to define a method trusting that it will be overridden later with a behavioural change, but
such change is visible only after the first time the setter is called. Moreover, this code would cache
the result instead of re-computing it every time. This can be very tricky in a
non-functional setting.
%\bruno{Yanlin please polish text/break long sentences.}
\end{comment}

@@

\begin{lstlisting}
interface Point{ ...//as before
  public static Point of(int _x, int _y){
    return new Point(){...
      public Point clone(){ return Point.super.clone();}};}}
\end{lstlisting}

### 2015-12-06 paper/sections/RelatedWork.tex
revised S7,8,9,10

\subsection{Multiple Inheritance in Object Oriented Languages}
Many authors have argued in favour or against multiple inheritance.  Multiple
inheritance is very expressive, but difficult to model and implement, and can
cause difficulty (including the famous diamond (fork-join)
problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.) in reasoning about 
programs. To conciliate the need for expressive power and simplicity, many
models have been proposed, including C++ virtual inheritance,
mixins~\cite{bracha90mixin}, traits~\cite{scharli03traits}, and hybrid model
such as CZ~\cite{malayeri2009cz}.  They provide novel programming architecture
models in the OO paradigm. In terms of restrictions set on these models, C++
virtual inheritance aims at a relative general model; mixin model add some
restrictions on the model; and trait model is the most restricted one (excluding
states, instantiation, etc).

@@

\paragraph{Mixins.}
Mixins are a more restricted model than C++ approach. Mixins allow to name
components that can be applied to various classes as reusable functionality
units. However, % they suffer from linearisation: the order of mixin application
% is relevant in often subtle and undesired ways. This hinders their usability
% and the ability of resolving conflicts:
the linearisation (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in some cases and
restricts the flexibility of mixin composition. To fight this limitation, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raises the complexity, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala traits are in fact more like linearised mixins.
Scala avoids the object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearised mixin composition. Other languages, such as 
Python, also use linearized mixins.
\begin{comment}
Python also offers multiple inheritance via linearised mixins. Indeed in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@ 
\end{comment}
\noindent Java interfaces and default methods does not use
linearisation: the semantics of Java \textbf{extends} clause in
interfaces is unordered and symmetric.

@@

\paragraph{CZ.}
Malayeri and Aldrich proposed a model CZ~\cite{malayeri2009cz} which aims to do
multiple inheritance without diamond problem. They divide inheritance into two
separate concepts: inheritance dependency and implementation inheritance. Using
a combination of \texttt{requires} and \texttt{extends}, a program with diamond
inheritance can be transformed to one without diamonds. Moreover, fields and
multiple inheritance can coexist. However as shown in the example in the
article, the untangling of inheritance also untangles class structure. In CZ,
not only the number of classes (from 4 to 6 in the stream example), but also the
class hirarchy complexity increases. \mixin does not complicate the
structure, and fields can also coexist with multiple inheritance.

@@

\paragraph{Traits vs Object Interfaces}
We consider object interfaces to be an alternative to traits or mixins.  In the
trait model two concepts (traits and classes) coexist and cooperate. Some
authors~\cite{BettiniDSS13} see this as good language design fostering good
software development by helping programmers to think about the structure of
their programs.  However, other authors see the need of two concepts and the
absence of state as drawbacks of this model~\cite{malayeri2009cz}.  Object
interfaces are units of reuse, and at the same time they provide factory methods
for instantiation and support state.  Our approach promotes the use of
interfaces instead of classes, in order to rely on the modular composition
offered by interfaces. Since Java was designed for classes, a direct classless
programming style is verbose and unnatural. However, annotation driven code
generation is enough to overcome this difficulty, and the resulting programming
style encourages modularity, composability and reusability, by keeping a strong
OO feel. In that sense, we promote object interfaces as being both units of
reusable code and object factories. Our practical experience is that, in Java,
separating the two notions leads to a lot of boilerplate code, and is quite
limiting when multiple inheritance with states is required.  Abstract state
operations avoid the key difficulties associated with multiple inheritance and
states, while still being quite expressive.  Moreover the ability to support
constructors adds additional expressivity, which is not available in approaches
such as Scala's traits/mixins.

@@

\item We provide \texttt{with-} methods as auxiliary constructors. A \texttt{with-} method is generated for each field, just like record update, returning the new object with that field updated
%. A \texttt{clone} method is generated for the interface, returning a copy of the current object.
Furthermore, we do automatic return type refinement for these kind of methods. This feature is comparatively useful in big examples, making operations and behaviours more flexible.%, which we will demonstrate later.
\item We provide two options for generating setters. There are two kind of setters which are commonly used, namely \textit{void setters} and \textit{fluent setters}. The only difference is that a fluent setter returns the object itself after setting, thus supporting a pipeline of such operations. The generation depends on which type of setter is declared in the interface by users.
\end{itemize}
\end{comment}

@@

In certain situations, object interfaces allow automatic refinement for \emph{return
types}. This is part of a bigger topic in class based languages: expressing and
preserving type recursion and (nominal/structural) subtyping at the same time.

@@

\subsection{Meta-programming Competes with Language Extensions}
The most obvious solution to adding features to a language is language
extension. It is often implemented as syntactic extensions that can be desugared
to the base language. For example, the Scala compiler was extended to directly
supports XML syntax. However, this approach does not support combining multiple
extensions into one. Scala-Virtualized is an extension to Scala which allows to
blend shallow and deep embedding of DSLs. It redefines Scala structures to
method calls, which can be overridden by DSL implementors. Therefore
Scala-Virtualized also reinterprets syntax. These extensions are admittedly
useful, however they are essentially creating a fork of the language, and rarely
the new fork gains enough traction to become the main language release. % On this
% topic we mention
SugarJ~\cite{erdweg11sugarj} is a Java-based extensible language allowing
programmers to extend it with custom features by definitions in meta-DSLs (SDF,
Stratego, etc). Using SugarJ requires learning various DSLs while our approach
is just using Java. However its intention is different from ours: one of the
purpose of SugarJ is to allow easy defition of sugars while our approach aims to
implement the desired program transformation in pure Java.

@@

Java-like languages tend to sit in the middle of two extremes: libraries
can not influence the type system, so many solutions valid in python or other dynamic
languages are not be applicable, or may be applicable at the cost of losing
safety.

Here compile time code generation comes at the rescue: if for a certain feature
(\mixin in our case) it is possible to use the original language syntax to
\emph{express-describe} any specific instantiation of such feature (annotating a
class and providing getters), then we can insert in the compilation process a tool
that examins and enrichs the code before compilation. No need to modify the
original source; for example we can work on temporary files. Java is a particular
good candidate for this kind of manipulation since it already provide ways to
define and integrate such tools in its own compilation process: in this way
there is no need of temporary files, and there is a well defined way of putting
multiple extensions together.

Other languages offer even stronger support to safe code manipulation:
Template Haskell~\cite{sheard2002template}, F\# (type providers)~\footnote{http://research.microsoft.com/fsharp/} and MetaFjig (Active Libraries)~\cite{servetto2010metafjig}
all allow to execute code at compile time and to generate code/classes that are transparently integrated in the program that is being generated/processed/compiled.
In particular, MetaFjig offers a property called \emph{meta-level-soundness}. In short this property ensures by construction that library code (even if wrong or non nonsensical) would never generate ill-typed code. This is roughly equivalent to 
\textit{Self coherence}, that we manually proof in the Appendix.
Since MetaFjig is not working on annotated classes, there is no direct equivalence on the overall theorem of safety we shown.

>>>>

\subsection{Multiple Inheritance in Object Oriented Languages}
Many authors have argued in favour or against multiple inheritance.  Multiple
inheritance is very expressive, but difficult to model and implement, and can
cause difficulty (including the famous diamond (fork-join)
problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.) in reasoning about 
programs. To conciliate the need for expressive power and simplicity, many
models have been proposed, including C++ virtual inheritance,
mixins~\cite{bracha90mixin}, traits~\cite{scharli03traits}, and hybrid model
such as CZ~\cite{malayeri2009cz}.  They provide novel programming architecture
models in the OO paradigm. In terms of restrictions set on these models, C++
virtual inheritance aims at a relative general model; mixin model adds some
restrictions on the model; and trait model is the most restricted one (excluding
states, instantiation, etc).

@@

\paragraph{Mixins.}
Mixins are a more restricted model than C++ approach. Mixins allow to name
components that can be applied to various classes as reusable functionality
units. However, % they suffer from linearisation: the order of mixin application
% is relevant in often subtle and undesired ways. This hinders their usability
% and the ability of resolving conflicts:
the linearisation (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in some cases and
restricts the flexibility of mixin composition. To fight this limitation, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raises the complexity, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala traits are in fact more like linearised mixins.
Scala avoids the object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearised mixin composition. Other languages, such as 
Python, also use linearized mixins.
\begin{comment}
Python also offers multiple inheritance via linearised mixins. Indeed in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@ 
\end{comment}
\noindent Java interfaces and default methods do not use
linearisation: the semantics of Java \textbf{extends} clause in
interfaces is unordered and symmetric.

@@

\paragraph{CZ.}
Malayeri and Aldrich proposed a model CZ~\cite{malayeri2009cz} which aims to do
multiple inheritance without diamond problem. They divide inheritance into two
separate concepts: inheritance dependency and implementation inheritance. Using
a combination of \texttt{requires} and \texttt{extends}, a program with diamond
inheritance can be transformed to one without diamonds. Moreover, fields and
multiple inheritance can coexist. However as shown in the example in the
article, the untangling of inheritance also untangles class structure. In CZ,
not only the number of classes (from 4 to 6 in the stream example), but also the
class hierarchy complexity increases. \mixin does not complicate the
structure, and fields can also coexist with multiple inheritance.

@@

\paragraph{Traits vs Object Interfaces}
We consider object interfaces to be an alternative to traits or mixins.  In the
trait model two concepts (traits and classes) coexist and cooperate. Some
authors~\cite{BettiniDSS13} see this as good language design fostering good
software development by helping programmers to think about the structure of
their programs.  However, other authors see the need of two concepts and the
absence of state as drawbacks of this model~\cite{malayeri2009cz}.  Object
interfaces are units of reuse, and at the same time they provide factory methods
for instantiation and support state.  Our approach promotes the use of
interfaces instead of classes, in order to rely on the modular composition
offered by interfaces. Since Java was designed for classes, a direct classless
programming style is verbose and unnatural. However, annotation-driven code
generation is enough to overcome this difficulty, and the resulting programming
style encourages modularity, composability and reusability, by keeping a strong
OO feel. In that sense, we promote object interfaces as being both units of
reusable code and object factories. Our practical experience is that, in Java,
separating the two notions leads to a lot of boilerplate code, and is quite
limiting when multiple inheritance with states is required.  Abstract state
operations avoid the key difficulties associated with multiple inheritance and
states, while still being quite expressive.  Moreover the ability to support
constructors adds additional expressivity, which is not available in approaches
such as Scala's traits/mixins.

@@

\item We provide \texttt{with-} methods as auxiliary constructors. A \texttt{with-} method is generated for each field, just like record update, returning the new object with that field updated
%. A \texttt{clone} method is generated for the interface, returning a copy of the current object.
Furthermore, we do automatic return type refinement for this kind of methods. This feature is comparatively useful in big examples, making operations and behaviours more flexible.%, which we will demonstrate later.
\item We provide two options for generating setters. There are two kinds of setters which are commonly used, namely \textit{void setters} and \textit{fluent setters}. The only difference is that a fluent setter returns the object itself after setting, thus supporting a pipeline of such operations. The generation depends on which type of setter is declared in the interface by users.
\end{itemize}
\end{comment}

@@

In certain situations, object interfaces allow automatic refinement for \emph{return
types}. This is part of a bigger topic in class-based languages: expressing and
preserving type recursion and (nominal/structural) subtyping at the same time.

@@

\subsection{Meta-programming Competes with Language Extensions}
The most obvious solution to adding features to a language is language
extension. It is often implemented as syntactic extensions that can be desugared
to the base language. For example, the Scala compiler was extended to directly
supports XML syntax. However, this approach does not support combining multiple
extensions into one. Scala-Virtualized is an extension to Scala which allows to
blend shallow and deep embedding of DSLs. It redefines Scala structures to
method calls, which can be overridden by DSL implementors. Therefore
Scala-Virtualized also reinterprets syntax. These extensions are admittedly
useful, however they are essentially creating a fork of the language, and rarely
the new fork gains enough traction to become the main language release. % On this
% topic we mention
SugarJ~\cite{erdweg11sugarj} is a Java-based extensible language allowing
programmers to extend it with custom features by definitions in meta-DSLs (SDF,
Stratego, etc). Using SugarJ requires learning various DSLs while our approach
is just using Java. However its intention is different from ours: one of the
purposes of SugarJ is to allow easy defition of sugars while our approach aims to
implement the desired program transformation in pure Java.

@@

Java-like languages tend to sit in the middle of two extremes: libraries
can not influence the type system, so many solutions valid in python or other dynamic
languages are not applicable, or have the cost of losing
safety.

Here compile time code generation comes at the rescue: if for a certain feature
(\mixin in our case) it is possible to use the original language syntax to
\emph{express-describe} any specific instantiation of such feature (annotating a
class and providing getters), then we can insert in the compilation process a tool
that examines and enriches the code before compilation. No need to modify the
original source; for example we can work on temporary files. Java is a particular
good candidate for this kind of manipulation since it already provide ways to
define and integrate such tools in its own compilation process: in this way
there is no need of temporary files, and there is a well-defined way of putting
multiple extensions together.

Other languages offer even stronger support to safe code manipulation:
Template Haskell~\cite{sheard2002template}, F\# (type providers)~\footnote{http://research.microsoft.com/fsharp/} and MetaFjig (Active Libraries)~\cite{servetto2010metafjig}
all allow to execute code at compile time and to generate code/classes that are transparently integrated in the program that is being generated/processed/compiled.
In particular, MetaFjig offers a property called \emph{meta-level-soundness}. In short this property ensures by construction that library code (even if wrong or unreasonable) would never generate ill-typed code. This is roughly equivalent to 
\textit{Self coherence}, that we manually proof in the Appendix.
Since MetaFjig is not working on annotated classes, there is no direct
equivalence on the overall theorem of the shown safety.
